---
title: ""
author: "Pavithra Govinda Raj"
date: "2026-01-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir= normalizePath('..'))
```

```{r load_project, include = FALSE}
library(ProjectTemplate)
load.project()
```


# Introduction 

This report applies Learning Analytics techniques to data from multiple runs of the MOOC “Cyber Security: Safety at Home, Online, and in Life”, developed by Newcastle University and delivered via the FutureLearn platform. The analysis follows the CRISP-DM (Cross-Industry Standard Process for Data Mining) cycle, progressing through business understanding, data understanding, data preparation, analysis, and evaluation to investigate learner engagement and attrition patterns.

The stakeholders are higher education institutions, course designers, learning technologists, academic researchers, and platform providers. Learner engagement is a critical challenge in MOOCs, where high dropout rates are common, and improving engagement can enhance learning outcomes, learner satisfaction, and the effective use of educational resources.

MOOC platforms generate rich interaction data through virtual learning environments, course content, and assessments, offering insights beyond traditional engagement metrics such as completion rates. When analysed within a structured framework like CRISP-DM, this data can reveal where learners disengage and which course design elements may contribute to attrition.

The study is guided by the research question:

> **"In which steps do users drop off the most?**"

The Learning Analytics process is considered successful if key points of disengagement are identified, or if learner attrition is shown to be evenly distributed across course steps, thereby informing evidence-based interventions and course improvements.

# CRISP-DM Cycle 1

## Business Understanding

The organisation that publishes the data is Newcastle University, in collaboration with the FutureLearn platform. The data is produced through the delivery of the MOOC “Cyber Security: Safety at Home, Online, and in Life” to learners from diverse geographic and educational backgrounds. The purpose of collecting and analysing this data is to document patterns of learner engagement and attrition, with the aim of informing course design decisions, improving learner retention, and enhancing the overall effectiveness of large-scale online education.

## Data Understanding

The data used in this analysis records learner activity metrics across seven weeks of the online course “Cyber Security: Safety at Home, Online, and in Life.” The dataset is sourced from the FutureLearn platform and captures learner interactions at multiple levels of granularity. For each week of the course, eight separate data files are provided, each representing a distinct aspect of learner engagement. As the data is generated directly by the learning platform, there are no general concerns regarding data quality. Further information on the structure and definitions of the recorded metrics is provided below.

- 

```{r} 
#EDA : Adoption Rate 
# Step 3: Create a bar graph of adoption rate per month
ggplot(adoption_pt_enrolments_7,
       aes(x = factor(month_enroled), y = adoption_rate)) +
  geom_bar(stat = "identity", fill = "#00C7C7") +
  labs(title = "Adoption Rate per Month",
       x = "Month",
       y = "Adoption Rate (%)") +
  scale_x_discrete(labels = c(
    "Jan", "Feb", "Mar", "Apr", "May", "Jun",
    "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
  )) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#completion rate
print(paste("Less than 1% completed the course:", total_completion_rate))
```



```{r}
#Identifying the drop off point
ggplot(drop_off_ordered, aes(x = week_step, y = number_of_learners)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Number of Learners at each Step",
    x = "Last Completed Step",
    y = "Number of Learners"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}

```


```{r}
#checking to see if the experience influences when a user leaves
cor(weekly_survey_responses$experience_rating,
    weekly_survey_responses$week_number,
    method = "spearman", use="complete.obs")
```

Positive sign → very slight tendency that higher experience ratings correspond to later leaving weeks
Magnitude is very close to 0 → practically no monotonic association
Essentially, experience rating and leaving week are independent in your dataset.

```{r}
#does the video duration influence last_completed_week number. interepretation, the users that completed the course have actually gone through all the videos and taken the time to complete them
cor.test(df$video_duration, df$last_completed_week_number, method="spearman")
```

As video_duration increases, last_completed_week_number always increases without exception.
Strength: rho = 1 → extremely strong positive monotonic relationship
Direction: positive → longer videos correspond to later leaving weeks
Significance: p < 2.2e-16 → statistically significant
Caution: ties reduce the precision of the exact p-value, but the conclusion of strong positive correlation is still valid.
In practice: rho = 1 with discrete steps often indicates that step number is perfectly aligned with cumulative video duration — which makes sense if video_duration is taken from the step the learner last completed.

```{r}
#how long does it take to complete one step
ggplot(summary_stats, aes(x=step, y=median_duration)) +
  geom_col() +
  labs(y="Median duration (secs)", x="Step")
```

Interpreting the plot:
The x-axis shows the steps in the course, probably in ascending order (1, 2, 3, ...).
The y-axis shows the median duration learners spent on each step, measured in seconds.
Each bar height corresponds to the median time learners took on that particular step.
From the plot, you can see:
Some steps have very low median durations (short bars), meaning learners spent less time on those steps on average.
Some steps have very high median durations (tall bars), meaning learners spent a longer time on those steps.
For example, there’s a very tall bar at around step 3, showing a large spike in median duration, indicating that step 3 took much longer on median compared to other steps.
Some steps show a cluster of short bars with very low durations, possibly indicating quick steps or skippable content.
The variability between steps in terms of median duration could suggest that some steps are more complex or require more effort.

```{r}
#time taken at eat step and the number of learners at each step
ggplot(step_summary, aes(x = step)) +
  geom_col(aes(y = median_duration_secs), fill = "steelblue", alpha = 0.7) +
  geom_line(aes(y = learner_count_scaled, group = 1),
            color = "firebrick", size = 1) +
  geom_point(aes(y = learner_count_scaled),
             color = "firebrick", size = 2) +
  scale_y_continuous(
    name = "Median Duration (seconds)",
    sec.axis = sec_axis(
      ~ . / max_duration * max_learners,
      name = "Number of Learners"
    )
  ) +
  labs(
    x = "Step",
    title = "Median Duration per Step with Learner Count Overlay"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Interpretation:
Median Duration per Step:
The height of the bars varies substantially between steps.
Some steps have very high median durations, such as step 3, where the median time exceeds 800 seconds (~13 minutes). This indicates that this step takes significantly longer for learners to complete than other steps.
Many steps have relatively short median durations below 200 seconds (~3 minutes), suggesting quicker completion.
The variation in median duration likely reflects different content types or difficulty levels across the steps.
Number of Learners per Step (Red line):
The red line shows a general declining trend in the number of learners as the step number increases.
Early steps have the highest learner counts, close to or above 1500 learners, indicating most learners start the course and engage with initial steps.
The number of learners drops steadily through the course, reflecting natural attrition or drop-off as learners progress.
There are some fluctuations where learner counts temporarily stabilize or increase slightly at some steps, possibly indicating re-engagement or course structure effects.
At some steps (e.g., around steps 18-19), the learner count drops sharply, suggesting significant dropout points.
Relationship between Duration and Learner Count:
Steps with longer median durations don’t always correspond to higher or lower learner counts.
For example, step 3 has a very high median duration but still retains a large number of learners.
In contrast, some steps with moderate or low median duration have low learner counts, which might indicate dropout is not necessarily driven by step length alone.
Overall conclusion:
Learners tend to spend different amounts of time on different steps, with some steps being notably longer.
Learner attrition is evident as the course progresses, with fewer learners reaching later steps.
The combined visualization helps identify critical steps that take longer and steps where learner drop-off is highest, useful for course design improvements.