---
title: "MAS8600_assignment"
author: "Pavithra Govinda Raj"
date: "2026-01-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction 

Learning Analytics has emerged as a pivotal area of focus within the field of Data Science, with the potential to significantly enhance the understanding of how students engage with educational content and environments. Defined as the "measurement, collection, analysis, and reporting of data about learners and their contexts," Learning Analytics plays a crucial role in optimizing learning experiences and outcomes. By leveraging diverse datasets, Learning Analytics aims to provide valuable insights into the effectiveness of educational methods, inform intervention strategies, and ultimately enhance student performance.

Traditionally, student engagement has been tracked through basic metrics such as attendance, which, while useful, fail to capture the full extent and quality of engagement. Beyond the classroom, various forms of data—ranging from interactions with Virtual Learning Environments (VLEs) and campus facilities to wellbeing referrals—are routinely collected. However, these datasets are often siloed and lack the integration needed to generate comprehensive insights.

This report focuses on a dataset collected from seven runs of the massive open online course (MOOC) titled "Cyber Security: Safety at Home, Online, and in Life," developed by Newcastle University and offered to the public via the FutureLearn platform. The data encompasses learner interactions throughout the course, along with profile-based characteristics, allowing for a deeper understanding of learner behavior and engagement. By aggregating and analyzing this data, this report seeks to identify patterns of engagement, determine factors that influence student success, and explore ways to enhance learning design and student interventions, especially for those at risk of underperformance.

Through the application of Learning Analytics, we aim to not only understand current learner engagement but also optimize future learning experiences, improving both student satisfaction and attainment.

# 
```{r}
#load('cached_objects.RData') - modify to specify for your objects
```


```{r} 
#EDA : Adoption Rate 
# Step 1: Calculate total learners across all months
total_learners <- sum(adoption_pt_enrolments_7$number_of_learners)

# Step 2: Calculate adoption rate for each month
adoption_pt_enrolments_7 <- adoption_pt_enrolments_7 %>%
  mutate(adoption_rate = (number_of_learners / total_learners) * 100)

# Step 3: Create a bar graph of adoption rate per month
ggplot(adoption_pt_enrolments_7, aes(x = factor(month), y = adoption_rate, fill = factor(month))) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  labs(title = "Adoption Rate per Month",
       x = "Month",
       y = "Adoption Rate (%)") +
  scale_x_discrete(labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun","Jul","Aug","Sep","Oct","Nov","Dec")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x labels for readability
```

```{r}
#completion rate
completed_course <- completed_pt_enrolments_7$number_of_learners[-nrow(completed_pt_enrolments_7)] |>
  na.omit() |>
  sum()

total_enroled <- adoption_pt_enrolments_7$number_of_learners |>
  na.omit() |>
  sum()

total_completion_rate <- (completed_course / total_enroled) * 100
print(paste("Less than 1% completed the course:", total_completion_rate))
```

```{r}
#Identifying the drop off point
ggplot(drop_off_ordered, aes(x = week_step, y = number_of_learners)) +
  geom_bar(stat = "identity", fill = "#007BA7") +
  labs(
    title = "Number of Learners at each Step",
    x = "Last Completed Step",
    y = "Number of Learners"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```
```{r}

```


```{r}
#checking to see if the experience influences when a user leaves
cor(weekly_survey_responses$experience_rating,
    weekly_survey_responses$week_number,
    method = "spearman", use="complete.obs")
```

Positive sign → very slight tendency that higher experience ratings correspond to later leaving weeks
Magnitude is very close to 0 → practically no monotonic association
Essentially, experience rating and leaving week are independent in your dataset.

```{r}
#does the video duration influence last_completed_week number. interepretation, the users that completed the course have actually gone through all the videos and taken the time to complete them
cor.test(df$video_duration, df$last_completed_week_number, method="spearman")
```

As video_duration increases, last_completed_week_number always increases without exception.
Strength: rho = 1 → extremely strong positive monotonic relationship
Direction: positive → longer videos correspond to later leaving weeks
Significance: p < 2.2e-16 → statistically significant
Caution: ties reduce the precision of the exact p-value, but the conclusion of strong positive correlation is still valid.
In practice: rho = 1 with discrete steps often indicates that step number is perfectly aligned with cumulative video duration — which makes sense if video_duration is taken from the step the learner last completed.

```{r}
#how long does it take to complete one step
ggplot(step_summary, aes(x=step, y=median_duration)) +
  geom_col() +
  labs(y="Median duration (secs)", x="Step")
```

Interpreting the plot:
The x-axis shows the steps in the course, probably in ascending order (1, 2, 3, ...).
The y-axis shows the median duration learners spent on each step, measured in seconds.
Each bar height corresponds to the median time learners took on that particular step.
From the plot, you can see:
Some steps have very low median durations (short bars), meaning learners spent less time on those steps on average.
Some steps have very high median durations (tall bars), meaning learners spent a longer time on those steps.
For example, there’s a very tall bar at around step 3, showing a large spike in median duration, indicating that step 3 took much longer on median compared to other steps.
Some steps show a cluster of short bars with very low durations, possibly indicating quick steps or skippable content.
The variability between steps in terms of median duration could suggest that some steps are more complex or require more effort.

```{r}
#time taken at eat step and the number of learners at each step
ggplot(step_summary, aes(x = factor(step))) +
  geom_col(aes(y = median_duration_secs), fill = "steelblue", alpha = 0.7) +
  geom_line(aes(y = learner_count_scaled, group = 1), color = "firebrick", size = 1) +
  geom_point(aes(y = learner_count_scaled), color = "firebrick", size = 2) +
  scale_y_continuous(
    name = "Median Duration (seconds)",
    sec.axis = sec_axis(~ . / max_duration * max_learners,
                        name = "Number of Learners")
  ) +
  labs(
    x = "Step",
    title = "Median Duration per Step with Learner Count Overlay"
  ) +
  theme_minimal()
```

